FraCaS: Temporal Analysis

### 2020-02-07 paper2020/paper.tex
complete description of temporal treatment

We adjust the montegovian semantics so that every category takes a
timespan as an additional parameter. This timespan serves as a time
reference for semantics, and we call it a \emph{temporal context}. This current interval propagates through the compositional
interpretation down to atoms. By default, every interpretation passes
the temporal context down to its components without changing
it. However some key elements will act on it on nontrivial ways, which
we proceed to detail below.

@@

In our representation we deal only with present and simple past
tenses.  Indeed we find that FraCas does not exercise additional
specific tenses (if a more complicated tense is used, the additional
information is also carried by adverbs or adverbial phrases, in a more
specific way.)

TODO: optional time context.

The procedure to handle tense annotations is as follows:
\begin{itemize}
\item If the temporal context was modified by an adverb (or an
  averbial clause, such as ``before James swam''), then the tense does
  not create a new interval, but it may constrain it. Typically, a
  past tense adds the constraint that the temporal context ends before
  the constant $now$, which is a constant representing the current
  time.
\item Otherwise, if the tense is the present, then the temporal
  context becomes the simple $(now,now)$ interval.
\item
  If the tense is the past, then we locally quantify over a time
  interval $[t_0,t_1]$, such that $t_1 < now$.
\end{itemize}

@@

\begin{enumerate}
\item[exact interval adverbs] For such adverbs, an exact interval is
  provided. Typical such adverbs specify in fact a single point in
  time (so the start and the end of the interval coincide).

@@

\item[existential interval] The majority of temporal adverbs
  existentially quantify over a timespan. Examples include ``since
  1991'', ``in 1996'', ``for two years'', etc. The common theme is to
  introduce the interval and then restrict its bounds or its duration
  in someway. Sometimes the restriction is an equality (``for exactly
  two hours'').

@@

\item[universal interval] A few adverbs introduce intervals via a
  universal quantification (sometimes with a constraint). Examples
  include: ``always'', ``never''.

@@

A common theme in the testsuite is to refer to previous occurences of
an event:

@@

To be able to conclude as the testsuite expects, we have to make sure
that the two occurences of ``Jones left'', in P1 and P3 refer to the
same time intervals.

@@

Indeed, following our account, the above inference problem is be
interperted as ∃x. novel(x) ∀t<NOW.  ∧ write(smith,x,t,t) [the scope
for the existential is extended according to our previous analysis of
anaphora]

However, the testsuite instructs that we should not be able to refute
280, with the justification that ``wrote a novel'' is a repeatable
accomplishment:

@@

∀t<NOW. ∃x. novel(x) ∧ write(smith,x,t,t)

Our analysis does not need to separate these as special cases. Indeed,
even if $write(smith,x)$ is an activity and thus subject to unicity of
action, in (280), $x$ is quantified existentially; we have two
different actions: $write(smith,x) t1 t2$ and $write(smith,y) t3 t4$,
and thus we can't deduce equality of the interval $t1,t2$ and $t3,t4$;
in turn the hypothesis can't be refuted.

@@

``start to'', ``finish'', etc.

example:

Definition finishVV : VVT := fun tvp => fun x => fun t0 t1 =>
exists t0', exists t1', t0 <= t1' <= t1 /\ tvp t0' t1' x.

>>>>

We adjust the montegovian semantics so that every category takes a
\emph{optional} timespan as an additional parameter. This timespan
serves as a time reference for semantics, and we call it a
\emph{temporal context}. This context propagates through the
compositional interpretation down to atoms. By default, every
interpretation passes the temporal context down to its components
without changing it. However some key elements will act on it on
nontrivial ways, which we proceed to detail below.

The timespan in the context is optional because, in certain
situations, the semantics is different depending on whether a timespan
has been specified externally or not, as we explain below.

@@

In our semantics we deal only with present and simple past
tenses.  Indeed we find that FraCas does not exercise additional
specific tenses (if a more complicated tense is used, the additional
information is also carried by adverbs or adverbial phrases, in a more
specific way.)

The procedure to handle tense annotations is as follows:
\begin{itemize}
\item If the temporal context was modified by an adverb (or an
  averbial clause, such as ``before James swam''), then the tense does
  not create a new interval, but it may constrain it. Typically, a
  past tense adds the constraint that the temporal context ends before
  the timepoint $now$, which is a logical constant representing the current
  time.
\item Otherwise, if the tense is the present, then the temporal
  context becomes the simple $(now,now)$ interval.
\item
  If the tense is the past, then we locally quantify over a time
  interval $[t_0,t_1]$, such that $t_1 < now$.
\end{itemize}

@@

\begin{enumerate}
\item[exact adverbs] For such adverbs, an exact interval is
  provided. Typical such adverbs specify in fact a single point in
  time (so the start and the end of the interval coincide).

@@

\item[existentially quantifying] The majority of temporal adverbs
  existentially quantify over a timespan. Examples include ``since
  1991'', ``in 1996'', ``for two years'', etc. The common theme is to
  introduce the interval and then restrict its bounds or its duration
  in someway. Sometimes the restriction is an equality (``for exactly
  two hours'').

@@

\item[universally quantifying] A few adverbs introduce intervals via a
  universal quantification (sometimes with a constraint). Examples
  include: ``always'', ``never''.

@@

A common theme in the testsuite is the reference to previous
occurences of an event:

@@

To be able to conclude as the testsuite expects, we have to make sure
that the two occurences of ``Jones left'' (in P1 and P3) refer to the
same time intervals.

@@

Indeed, following our account, the above (contradictory) inference problem is to be
interperted as
∃x. novel(x) ∧
∃t1,t2. [t1,t2] ⊆ 1991 ∧ write(smith,x,t1,t2) ∧
∃t3,t4. [t3,t4] ⊆ 1992 ∧ write(smith,x,t3,t4)
⟶ ⊥

Thanks to unicity of action we find [t1,t2] = [t3,t4], and due to 1991
and 1992 being disjoint we obtain contradiction.

[the scope
for the existential is extended according to our previous analysis of
anaphora]

However, the testsuite instructs that we should \emph{not} be able to refute
280, with the justification that ``wrote a novel'' is a repeatable
accomplishment:

@@

Here our interpretation is:

∃x. novel(x) ∧
∃t1,t2. [t1,t2] ⊆ 1991 ∧ write(smith,x,t1,t2) ∧
∃y. novel(y) ∧
∃t3,t4. [t3,t4] ⊆ 1992 ∧ write(smith,x,t3,t4)
⟶ ⊥

Our analysis does not need to treat this last case specially. Indeed,
even if $write(smith,x,.,.)$ is an activity and thus subject to
unicity of action, in (280), $x$ is quantified existentially; we have
two \emph{different} actions: $write(smith,x,t1,t2)$ and
$write(smith,y,t3,t4)$, and thus we can't deduce equality of the
interval $t1,t2$ and $t3,t4$; in turn the hypothesis can't be refuted.

@@

The final class of lexeme carrying a temporal-dependent semantics are
verbs taking a proposition as argument, like ``finish'', ``start'',
etc. These verbs modify the temporal in non-trivial ways. Consider for
example ``finish to ...''. The timespan of the argument of finish
should end within the timespan of the finishing action:

⟦finish to s⟧(t0,t1) = ∃(t0',t1'). t1' ∈ [t0,t1] ∧ s(t0',t1')
